\section{Morse Functions}

\begin{definition}[Critical Point]
    Let $M$ be a manifold and $f: M \to \R$ be a smooth function. A critical point of
    $f$ is a point $p \in M$, such that $df(p) = 0$.
\end{definition}

\begin{remark}
    If $M$ is compact, $f$ always has critical points.
\end{remark}

\begin{definition}[Poisson Bracket]
    Let $X$ and $Y$ be vector fields on a manifold $M$. Then for a point $p \in M$ define
    the Poisson Bracket by
    \[  [X, Y]f := XYf - YXf \]
    The Poisson Bracket is a Lie Bracket on the $\R$-vectorspace of vector fields.
    Also, if $p$ is a critical point of a function $f$, then $[X, Y]f (p) = 0$.
\end{definition}

\begin{definition}[Hessian]
    In $\R^n$, we are used to defining the Hessian of a function, that is the second order
    derivative. On a manifold, this is not so easy, since the second order derivative will 
    always depend on the local coordinates. We will have to do with defining 
    the second order derivative just on the critical points of a function:
    For $p \in M$ a critical point of a function $f$, $x, y \in T_p(M)$, choose $X, Y$
    vectorfields extending $x$ and $y$ locally, i.e. with $Y(p) = y$ and $X(p) = x$. 
    Then define
    \[ \opd^2 f (x, y) (p) = XYf (p) = xYf(p)\]
\end{definition}

\begin{lemma}
    $d^2 f (\cdot, \cdot) p$ is a symmetrical bilinear form if $p$ is a critical point of
    $f$.
\end{lemma}

\begin{proof}
    Note that
    \[ d^2 f (x, y) (p) - d^2 f (y, x) (p) = [X, Y]f (p) = 0  \]
    so $XYf(p)$ is symmetrical at any critical point and
    \[ xYf (p) = XYf (p) \]
    for any choice of $X$ and $Y$, so $d^2 f(\cdot, \cdot) (p)$ does not depend
    on $X$, and by the same argument because it is symmetrical it does not depend on $Y$.
    Then $d^2 f (x, y) (p)$ is a well defined, symmetrical bilinear form.
\end{proof}

\begin{definition}[non-degeneracy, index]
    We call a critical point of a function $f: M \to \R$ \textit{non-degenerate}, if
    the bilinear form $d^f(\cdot, \cdot)(p)$ is non-degenerate. \\
    We define the index of $p$ as the maximal dimension of negative definite subspaces,
    on which $d^2 f (\cdot, \cdot) (p)$ is negative-definite.\\
    Note that with local coordinates $(x_1, ..., x_n)$ around $p$, we get an induced Basis 
    of the tangent space $T_p(M)$ as $B = (\pderive{x^1}, ..., \pderive{x^n})$, and then
    if $\Phi_B$ is the coordinate isomorphism, we get a Matrix $A$, such that
    \[ d^2 f (x, y) (p)  = \varphi_B (x)^T A \varphi_B(y) \]
    $A$ can be given by 
    \[ A = \left(\pdderive[f]{x_i}{x_j}(p)\right)_{1 \leq i, j \leq n} \]
    Then the index of $p$ is the number of ngative eingenvalues of $A$.
\end{definition}

\begin{definition}
    A smooth function $f: M \to \R$ is called a \textit{Morse Function}, if all its critical
    points are non-degenerate.
\end{definition}

\begin{lemma}
    Let $M \subseteq \R^n$ be a submanifold. Then for almost every point $p /in \R^n$,
    the function
    \begin{align*}
        f: M & \to \R \\
        q & \mapsto || q - p ||^2
    \end{align*}
    is a Morse Function.
\end{lemma}

\begin{proof}
    % TODO
\end{proof}

\begin{remark}
    Note that by Whitney's embedding theorem, there exist Morse-Functions on any Manifold.
\end{remark}

\begin{theorem}
    Any smooth function $f: M \to \R$ and all its derivatives can be uniformly approximated
    by a Morse Function.
\end{theorem}

\begin{proof}
    % TODO
\end{proof}

\begin{theorem}
    Let $M$ be a compact manifold. Then the set of Morse functions is dense in 
    $\mathcal{C}^{\infty}(M)$.
\end{theorem}

\begin{proof}
    %TODO
\end{proof}

\begin{theorem}[Morse Lemma]
    Let $f: M \to \R$ be a smooth function. Let $p$ be a non-degenerate critical point 
    of index $k$ of $f$. Then there exist local coordinates $(x_1, ..., x_n)$ in a 
    neighburhood $U$ of $p$, such that
    \[ f = f(p) - x_1^2 - ... - x_k^2 + x_{k + 1}^2 + ... + x_n^2 \]
    and
    \[ x_1(p) = ... = x_n(p) = 0 \]
    $U$ is called \textit{Morse chart}.
\end{theorem}

\begin{proof}
    %TODO
\end{proof}

\begin{corollary}
    Non-degenerate critical points are isolated.
\end{corollary}

\begin{remark}
    A critical point of Index $k$ of f is a critical point of index $n-k$ of $-f$.
\end{remark}
